Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-10 paper/type-inference.tex

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints, which are
subsequently solved enmasse. The motivation behind the whole-program
approach to constraint generation is twofold: it simplifies
elaboration functions and makes presentation easier, and second, it
naturally generalizes to mutual recursion. Nonetheless, we do not
intend our type inference to be a whole-program analysis for (a). it
preempts opportunities for separate compilation and dynamic linking,
and (b). it is expensive and an overkill in most practical cases. We
therefore reclaim the compositionality of type inference by solving
the constraints in a compositional fashion. In more practical terms
this means that our constraint solving algorithm visits and solves
every constraint (or, every set of mutually recursive constraints)
only once. It composes computed solutions to solve other constraints
that depend on the solved constraints. Importantly, the failure to
solve a dependent constraint does not result in backtracking. We now
describe our compositional constraint solving algorithm in detail.

\paragraph{Notation and terminology} In a validity constraint, a
predicate variable occuring on the left side of the turnstile is said
to occur negatively. In contrast, a predicate variable occuring on the
right side is said to occur positively.

Given a set of constraints, we first build a dependency graph ($G_c$)
with validity constraints as nodes, and dependencies between them
captured as edges. In particular, there exists an edge from a
constraint $c_2$ to a constraint $c_1$ in the graph if any of the
following conditions hold:
\begin{itemize}
\item $c_2$ and $c_1$ have atleast one common predicate variable
($\varphi$) that occurs positively in $c_2$ and negatively in $c_1$.
\item The set of predicate variables that occur negatively in $c_2$ is
a (non-strict) subset of the corresponding set for $c_1$.
\end{itemize}
The first condition intuitively corresponds to a case, where
expression or type, whose region elaboration is constrained by $c_2$
refers to a method or a class, whose unknown precondition is
constrained by $c_1$. The common predicate variable ($\varphi$)
represents the unknown precondition in this case. The dependency from
$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
before $c_2$ is solved, thus enforcing the rule that the precondition
of a method must not depend on its calling context. Since mutually
recursive definitions refer each other, the first condition also
results in bidirectional dependencies between mutually recursive
constraints (and self-loops over recursive constraints). The second
condition captures two kinds of dependencies. First, the dependency
from the constraints over a method precondition to the constraints
over the precondition of the class containing the method. This
captures our preference that the constraints over a class's region
parameters should not depend on the idiosyncracies of its methods. Any
additional constraints required by any of its methods must be captured
in the precondition of the method itself (well-formedness rules allow
this possibility). The second condition also adds bidirectional
dependencies between two validity constraints over same set of
predicate variables, thus requiring them to be solved simultaneously.

Next, we convert the dependency graph over constraints into a
dependency DAG ($G_C$) over sets of constraints, where each set
represents a strongly connected component in the dependency graph.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph. Accordingly, we
obtain a topological ordering of nodes in the graph $G_{{C}}$, and
solve the sets of constraints in that order. The solutions obtained
after solving a constraint set are applied to the constraints in
subsequent sets before attempting to solve them.

>>>>

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints, which are
subsequently solved enmasse. The motivation behind the whole-program
approach to constraint generation is twofold: it simplifies
elaboration functions and makes presentation easier, and second, it
naturally generalizes to mutual recursion. Nonetheless, we do not
intend our type inference to be a whole-program analysis for (a). it
preempts opportunities for separate compilation and dynamic linking,
and (b). it is expensive and an overkill in most practical cases. We
therefore reclaim the compositionality of type inference by solving
the constraints in a compositional fashion. In more practical terms
this means that our constraint solving algorithm visits and solves
every constraint (or, every set of mutually dependent constraints)
only once. It composes computed solutions to solve other constraints
that depend on the solved constraints. Importantly, the failure to
solve a dependent constraint does not result in backtracking. We now
describe our compositional constraint solving algorithm in detail. To
simplify the presentation of our algorithm, we assume that are no
mutually recursive definitions in the source program. Recursive
definitions are nonetheless allowed.

\paragraph{Terminology} In a validity constraint, a predicate variable
occuring on the left side of the turnstile is said to occur
negatively, or with \emph{negative polarity}. In contrast, a predicate
variable occuring on the right side is said to occur positively, or
with \emph{positive polarity}. A validity constraint
\emph{constrains} the set of predicate variables that occur
negatively in the constraint, while it \emph{uses} the set of
predicate variables that occur positively. A constraint is said to be
\emph{recursive} if it constrains and uses a predicate variable.
% A pair of validity
% constraints are said to be \emph{mutually dependent} if there exists a
% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
% with opposite polarities in both the constraints. In such case, the
% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
% be mutually dependent. We call the transitive closure of mutual
% dependency relation as \emph{transitive dependency} relation. Like
% mutual dependency, transitive dependency is also extended to predicate
% variables.

Given a set of validity constraints, we first build a dependency graph
($G_c$) with constraints as nodes, and dependencies between them
captured as edges. There exists an edge from a constraint $c_2$ to a
constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
any of the following conditions hold:
\begin{itemize}
\item $c_1$ constrains a predicate variable that $c_2$ uses. 
\item $c_1$ constrains a (non-strict) subset of predicate variables
that $c_2$ constraint.
\end{itemize}
The first condition intuitively corresponds to a case, where
expression or type, whose region elaboration is constrained by $c_2$
refers to a method or a class, whose unknown precondition is
constrained by $c_1$. The common predicate variable ($\varphi$)
represents the unknown precondition in this case. The dependency from
$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
before $c_2$ is solved, thus enforcing the rule that the precondition
of a method must not depend on its calling context. 
% Note that the first condition results in self-loops over recursive
% constraints in the dependency graph.  Since mutually recursive
% definitions refer each other, the first condition also results in
% bidirectional dependencies between mutually recursive constraints
% (and self-loops over recursive constraints). 
The second condition captures two kinds of dependencies. First, the
dependency from the constraints over a method precondition to the
constraints over the precondition of the class containing the method.
This captures our preference that the constraints over a class's
region parameters should not depend on the idiosyncracies of its
methods. Any additional constraints required by any of its methods
must be captured in the precondition of the method itself
(well-formedness rules allow this possibility). The second condition
adds bidirectional dependencies between validity constraints that
constraint the same set of predicate variables.

Next, we convert the dependency graph over constraints into a
dependency DAG ($G_C$) over sets of constraints, where each set
represents a strongly connected component in the dependency graph.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph. Accordingly, we
obtain a topological ordering of nodes in the graph $G_{{C}}$, and
solve the sets of constraints in that order. The solutions obtained
after solving a constraint set are applied to the constraints in
subsequent sets before attempting to solve them. Consequently, when
the turn of a constraint set ($C$) arrives during the constraint solving
process, it satisfies certain properties:
\begin{itemize}
\item There exists only one predicate variable that is either
constrained or used by the constraints in the set ($C$). The variable is
called set's \emph{subject}. This property follows from (a). the fact
that all the dependency constraints have already been solved (and
solutions applied), and (b). the assumption that there are no mutually
recursive definitions. 
\item All the constraints that constrain the set's subject are present
in the set. This follows from our definition of the dependency relation.
\end{itemize}

Solving the set ($C$) of constraints entails finding an assignment for
$C$'s subject, and also all the region variables that occur free in
$C$, such that the solutions satisfy the well-formedness constraints.
We now formalize the constraint satisfaction problem, and its
solution.

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem})}
A constraint satisfaction problem is a tuple
$(C,\rhoenv,\varphi,\bar{\rho} \in \bar{\rhoenv})$, where $C$, the set
of validity constraints, is of the form:
\begin{center}
\(
  \bigcup_{i=1}^n \{
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i} \conj F_i(\varphi)}
  \}
\)
\end{center}
The subject of the set is $\varphi$, and its free region variables
range over $\rhobar$. The solution to the constraint satisfaction
problem is a pair $(\substFn,\phisol)$, where $\substFn$ is a map from
region variables to region identifiers such that $\substFn(\rho_j) \in
\rhoenv_j$, for every $j$, and $\phisol$ is a constraint formula such
that:
\begin{itemize}
\item $\phisol$ is well-formed under $\rhoenv$ (i.e.,
$\tywf{\rhoenv}{\phisol}$).
\item $\phisol$ satisfies every validity constraint in $\substFn(C)$.
That is, the following sequent is valid for every $i$:
\begin{center}
\(
  \isvalid{\substFn(\phictxt^{i}) \conj \phisol}
          {\substFn(\phicstr^{i}) \conj (\substFn \circ F_i)(\phisol))}
\)
\end{center}
\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
that $\phisol'$ satisfies every validity constraint in $\substFn(C)$,
and $\phisol'$ is strictly weaker than $\phisol$ (i.e.,
$\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
\end{itemize}
\end{definition}
